\documentclass[11pt, preprint]{aastex}

\include{computational_defs}

\begin{document}

\title{\bf Computational Physics Project / Telescope Diffraction Limit}

This project involves a calculation of the diffraction pattern due to
the finite aperture of a telescope.

If a plane wave enters a well-focused telescope along its axis of
symmetry (on-axis), then geometric optics tells us that all of the
light will be focused at a single point. However, because light is a
wave, in fact it will form a finite-sized image at the focus. The size
and shape of the image is related to the aperture of the telescope.

Specifically, the image formed (the point spread function, or PSF) can
be shown to be the squared amplitude of the Fourier transform of the
complex pupil function:
\begin{equation}
P(x,y) = P_r(x, y) \exp\left(-i 2\pi W(x, y) / \lambda\right)
\end{equation}
where $P_r$ is a function (between 0 and 1) of position on the
aperture and $W$ is a phase shift (induced, for example, by
imperfections in the shape of the mirror, or by the
atmosphere). $\lambda$ is the wavelength of the light wave.

More specifically, the PSF in the focal plane coordinate system of
$x_f$ and $y_f$ can be written:
\begin{equation}
{\rm PSF}(x_f, y_f) = {\rm FT}\left[P(x, y)\right]\left(k_x=
\frac{x_f}{\lambda f}, k_y = \frac{y_f}{\lambda f}\right),
\end{equation}
where $f$ is the focal length. Note that the relationship between
incoming angular scale $\Delta\theta$ and displacement in the focal
plane $\Delta x_f$ is
\begin{equation}
\frac{\Delta x}{\Delta\theta} = f
\end{equation}
so $k_x$ in the Fourier Transform is evaluated at $\theta_x /
\lambda$, and same in $y$. We call it $\theta$ because it corresponds
to the incoming angle on the sky.

In the simplest case, $P(r, \phi)$ is just $1$ within some radius
$D/2$, and $0$ outside.  In this case, the image formed is:
\begin{equation}
I = C\left(\frac{J_1(\pi r_f)}{\pi r_f}\right)^2
\end{equation}
where $J_1$ is the first-order Bessel function, $r_f=
D\theta/\lambda$, and $\theta$ is radial component of the input angle
of light into the telescope.  This is known as the Airy pattern.

\section{Prep work}

\begin{itemize}
\item $J_1$ can be calculated as:
  \begin{equation}
   J_1 = \frac{1}{\pi} \int_0^\pi \dd\theta\, \cos\left(\theta -
   x\sin\theta\right)
  \end{equation}
  You should in general use numerical libraries to calculate it.  But
  this one time use Simpson's rule to check the NumPy implementation
  of the calculation. Note that NumPy (and other libraries) use a very
  accurate analytic approximation to evaluate this function, not an
  explicit integral.
\item Many reflective telescopes have a Cassegrain design
  (\href{https://en.wikipedia.org/wiki/Cassegrain_reflector}{\color{blue}
  see Wikipedia}), which causes the aperture to be a donut, with a
  hole in the middle. Can you write analytically (in terms of $J_1$)
  what you expect the PSF to be? Plot the result.
\item We will want to simulate small imperfections in the phases from
  different causes. To do so it will be useful to be able to create
  {\it Gaussian random fields}. These are random fields whose values
  have Gaussian distributions, and have random ``phases.''
  Specifically, they can be created by choosing independent Gaussian
  random values for Fourier modes in $\vec{k}$-space, and Fourier
  transforming back to configuration space. The random distributions
  have mean zero and a variance that is a function of the scalar $k$,
  called $P(k)$ or the ``power spectrum.'' First, create a function
  that returns a set of amplitudes $a(\vec{k})$ for each mode. The
  phase differences will be real, but the Fourier amplitudes will not
  be. You must set the amplitudes with the right symmetries to
  guarantee that their Fourier transform will be real!
\item Now create a function that will use the function from the
  previous step plus an FFT to produce a random
  field with some user-specified $P(k)$. You may consult
  \href{http://andrewwalker.github.io/statefultransitions/post/gaussian-fields/}{\color{blue}
    the example code here}. Try using
  \begin{equation}
    P(k) \propto k^{n} \exp\left(- k^2 / k_c^2\right)
  \end{equation}
    for several values of $n<0$ and $k_c$. Plot the random fields
    produced and comment on the differences.
\end{itemize}

\section{Calculating Ideal PSFs}

\begin{itemize}
\item Perform a numerical FFT of a simple circular aperture, in order
  to verify the above result in detail. Note that a discrete FFT is
  effectively a FFT of a periodically repeated pattern, which is {\it
    not} what you want. To simulate an isolated aperture, you should
  make the size of the grid at least twice the size of the aperture
  itself.
\item Compare the residuals between your calculation and the analytic
  calculation, for different choices of the grid resolution and size
  relative to the aperture.
\item Show how the size of the PSF varies with aperture size.
\item For a Cassegraine telescope, to hold the secondary mirror in
  place there need to be some structural elements.  Often these take
  the form of several struts holding it in place, which cause
  obscuration of the aperture (basically, small lines with $P=0$
  aligned radially). Look up some images of stars on the web from the
  Hubble Space Telescope, guess how many such struts it has, and try
  to reproduce the diffraction spikes (roughly) using your model. Do
  you notice anything about the patterns of color in the diffraction
  spikes that you can explain with your model?
\end{itemize}

\section{Calculating PSFs in Imperfect Systems}

\begin{itemize} 
\item Small imperfections in the mirror cause phase shifts, because
  they change the path length of the light in a way that depends on
  location in the aperture. Simulate such shifts in your model by
  creating a Gaussian random field for the phase offsets $W$ (which is
  in the same units as $\lambda$) in each pixel. Use a $P(k)$ with
  power at large $k$. Under what conditions do you expect these shifts
  to become a major problem? Can you verify that?
\item For a ground-based telescope, the atmosphere causes the light
  coming into the telescope to not be a plane wave. Instead it is
  ``wrinkled'' with a coherence length of about 20 cm. Create a
  Gaussian random field whose power spectrum cuts off on smaller
  scales, and above those scales has fluctuations substantially larger
  than the wavelength of the light. The random field should look
  ``smooth'' on small scales.  How do these errors affect the point
  spread function?
\item Simulate a long exposure by combining (in intensity) the PSFs
  from many different realizations of the Gaussian random
  field. Compare the individual frames to the total.
\end{itemize}

\end{document}
